Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-03-25 paper/zhang-cook-laa-langrene-menendez.Rmd
clean up figures

knitr::read_chunk(
  here::here("scripts/implementation.R")
)

@@

```{r toy-pca, fig.asp=1, fig.height=2.5, fig.cap = "Search paths of CRS (green) and PD (brown) in the PCA-reduced basis space for 1D projection problem on the five-variable dataset, \\code{boa5} using holes index. The basis space, a 5D unit sphere, is projected onto a 2D circle by PCA. All the bases in PD has been flipped for easier comparison of the final bases and a grey dashed line has been annotated to indicate the symmetry of the two start bases."}
```

@@

<!-- something about the diagnostic plots  -->
Four diagnostic plots have been introduced to investigate the progression of the optimiser and its coverage in search space. The implementation of these visualisations is designed to be easy-to-use, each plot can be produced by supplying the data object to the corresponding function. More advanced users may decide to modify these basic graphs using the provided wrapper functions, or even build their own.

<!-- summarise what we have done and what might be done in the future. -->
Most of the work in this project has been translated into code in two packages: the collection of the data object is implemented in the existing \pkg{tourr} [@tourr] package; while the manipulation and visualisation of the data object are implemented in the new \pkg{ferrn} package. Equipped with handy tools to diagnose the performance of optimisers, future work can extend the diagnostics to a wider range of index functions, i.e. scagnostics, association, and information index [@laa2020using] and understand how the optimisers behave for index functions with different structures.

>>>>

@@

```{r toy-pca, fig.height = 5, fig.cap = "Search paths of CRS (green) and PD (brown) in the PCA-reduced basis space for 1D projection problem on the five-variable dataset, \\code{boa5} using holes index. The basis space, a 5D unit sphere, is projected onto a 2D circle by PCA. All the bases in PD has been flipped for easier comparison of the final bases and a grey dashed line has been annotated to indicate the symmetry of the two start bases."}
```

@@

This work might be considered an effort to bring transparency into algorithms. Although little attention is paid by algorithm developers to providing ways to output information during intermediate steps, this is an important component for enabling others to understand and diagnose the performance. Algorithms are an essential component of artificial intelligence that is used to make daily life easier. Interpretability of algorithms is important to guard against aspects like bias and inappropriate use. The data object underlying the visual diagnostics here is an example of what might be useful in algorithm development generally.

### 2021-03-25 paper/zhang-cook-laa-langrene-menendez.tex
clean up figures

The focus of this paper is on the optimisation problem arising in the
projection pursuit guided tour \citep{buja2005computational}, an
exploratory data analysis tool used for detecting interesting structures
in high-dimensional data through a set of lower-dimensional projections
\citep{cook2008grand}. The goal of the optimisation is to identify the
projection, represented by the projection matrix, that gives the most
interesting low-dimensional view. A view is interesting if it can show
some structures of the data other than a normal distribution, e.g.,
bi-modality, clustering, and outliers.

The optimization challenges encountered in the projection pursuit guided
tour problem are common to those of optimization in general. Examples
include the existence of multiple optima (local and global), the
trade-off between computational burden and proximity to the optima, or
dealing with noisy objective functions that might be non-smooth and
non-differentiable \citep{jones1998efficient}. The visualization tools
and optimization methods presented in this paper can therefore easily be
applied to other optimization problems.

The remainder of the paper is organised as follows. Section \ref{optim}
provides an overview of optimisation methods, specifically line search
methods. Section \ref{tour} reviews projection pursuit guided tour,
defines the optimisation problem and outlines three existing algorithms.
Section \ref{vis-diag} presents the new visual diagnostics. A data
structure is defined to capture information during the optimisation, and
different diagnostic plots are designed with the data collected. Section
\ref{application} shows applications of how these diagnostic plots can
be used to examine the performance of different optimisers and guide
improvements to the existing algorithms. Finally, Section
\ref{implementation} describes the R package: \CRANpkg{ferrn}
\citep{ferrn}, that implements the visual diagnostics.

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-pca-1}

@@

This paper has provided several visual diagnostics that can be used for
diagnosing a complex optimisation procedure. The methods were
illustrated using the optimisers available for projection pursuit guided
tour. The methods are implemented in the package \pkg{ferrn}, and
changes were made in the \pkg{tourr} package to generate the necessary
data. This work complements with the work of \citet{laa2020using} and
will be useful for researchers when designing new projection pursuit
indexes.

>>>>

The focus of this paper is on the optimisation problem arising in the
projection pursuit guided tour \citep{buja2005computational}, an
exploratory data analysis technique used for detecting interesting
structures in high-dimensional data through a set of lower-dimensional
projections \citep{cook2008grand}. The goal of the optimisation is to
identify the projection, represented by the projection matrix, that
gives the most interesting low-dimensional view. A view is interesting
if it can show some structures of the data that depart from normalality,
such as bimodality, clustering, or outliers.

The optimization challenges encountered in the projection pursuit guided
tour problem are common to those of optimization in general. Examples
include the existence of multiple optima (local and global), the
trade-off between computational burden and proximity to the optima, or
dealing with noisy objective functions that might be non-smooth and
non-differentiable \citep{jones1998efficient}. The visualization tools,
optimization methods and conceptual framework presented in this paper
can therefore be applied to other optimization problems.

The remainder of the paper is organised as follows. The next section
provides an overview of optimisation methods, specifically line search
methods. A review of the projection pursuit guided tour, an overview of
the optimisation problem and outlines of three existing algorithms
follows. The third section presents the new visual diagnostics,
including the the design of a data structure to capture information
during the optimisation, from which several diagnostic plots are
created. An illustration of how the diagnostic plots can be used to
examine the performance of different optimisers and guide improvements
to the existing algorithms applications is shown using simulated data.
Finally, an explanation of the implementation in the R package,
\CRANpkg{ferrn} \citep{ferrn}, is provided.

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-pca}

@@

This paper has provided several visual diagnostics that can be used for
diagnosing a complex optimisation procedure. The methods were
illustrated using the optimisers available for projection pursuit guided
tour. Here the constraint is the orthonormality condition of the
projection bases. The approach used broadly applies to other constrained
optimisers.
